Transactional Memory Concurrency Verification with Landslide

### 2018-03-16 htm/paper.tex
proofreading pass

\begin{abstract}
Hardware transactional memory is a recently-introduced concurrent programming paradigm
%backed by a set of new instructions on recent x86 CPUs,
which allows programmers to elide locks for performance in low-contention workloads.
However,
%using this feature
it
comes at a cost in implementation complexity:
%programmers must support their
fast-path code must be accompanied by backup paths to handle transaction failure.

We extend Landslide, a popular stateless model checker,
with a concurrency model for transactional memory
and evaluate it on several real-world transactional
benchmarks and data structure implementations.
\end{abstract}

@@

Transactional Synchronization Extensions (TSX) \cite{transactional-memory}
is an instruction set extension for x86 CPUs which adds hardware-based transactional memory.
The processor uses its existing cache coherence algorithm to
check for memory conflicts with other cores while temporarily staging a sequence of memory accesses.
If no other CPU accesses the same memory during the transaction,
the access sequence is committed to main memory atomically (with respect to visibility to other CPUs).
Otherwise, the accesses are discarded, the CPU's local state is reverted, and the transaction returns a failure code.

This feature can be used to replace conventional locking in performance-critical concurrent programs.
When a concurrent workload accesses largely thread-local data,
or disjoint sections of a shared data structure,
the contention rate between threads is low,
and transactions will often succeed.
Compared to programs which use conventional locks,
which use bus-locking atomic accesses even in the fastest code path,
TSX provides substantial performance improvements in such programs \cite{htm-experience, htm-performance, htm-mario}.
However, the possibility of transaction failure introduces additional implementation complexity:
programmers must also provide a backup plan
to safely resolve contention between threads,
usually involving conventional synchronization.
These backup paths must coordinate not only with other backup paths
but also with other fast paths which another thread may begin after the original transaction failed,
which even in the simplest transactions requires complex synchronization sequences
\cite{htm-mario}.
This introduces an additional dimension of nondeterminism into an already concurrent program,
and moreover, because transaction failure is expected to be rare,
% だからこそ
obscure interleavings between failure paths are difficult to expose during testing.

@@

{\bf Hardware transactional memory.}
TSX was supported on consumer hardware for the first time by Intel's Haswell architecture \cite{htm-haswell},
which extends the x86 instruction set to provide
{\tt xbegin}, {\tt xend}, and {\tt xabort} for beginning, committing, and aborting
transactions, respectively.
Higher-level programming languages or compilers may offer libraries or intrinsics to access these instructions;
for C and C++ GCC provides intrinsics named {\tt \_xbegin()} and so on \cite{htm-gcc}.
Figure~\ref{fig:htm-example} shows an example program using TSX to synchronize access to a shared counter,
including a failure path which defaults to a conventional lock.
This example actually has a bug, which we will discuss in the next section;
the reader is encouraged to try to spot it before then.
Several related works formally prove the correctness of transactional memory {\em implementations}
\cite{specifying-verifying-tm,tm-correctness,tm-completeness,mc-tm-with-spin},
but verifying the client programs written to use transactions remains an open problem.

@@

@@

Under software transactional memory (STM) \cite{stm-pldi06},
memory conflicts with other threads are the only reason for transaction failure
(apart from programmer-supplied explicit aborts);
hence, depending on program semantics, some transactions may be guaranteed to succeed.
However, hardware implementations (HTM) may also fail transactions
for several other reasons such as random system interrupts or exhausting the CPU's cache capacity.
Because timer interrupts can in principle occur at any moment,
and with arbitrary frequency (observable by the program, perhaps as a result of a heavily-loaded system),
in this paper we will simplify the failure model by saying that HTM transactions can fail for any reason.
We defer discussion of programs which distinguish the reason for aborts through the failure code to Section~\ref{sec:warpzone}.

{\bf Stateless model checking.}
Model checking (MC) \cite{verisoft} is a testing technique for systematically executing and verifying
the possible thread interleavings of a concurrent program.
The main research challenge is to cope with exponential explosion of the state space,
which is $O(n^k)$ for a program with $n$ operations and $k$ threads.
Some {\em stateful} MCs explicitly store and compare all visited states of the program being tested \cite{spin},
which both keeps track of test coverage
and allows identifying identical states to avoid testing redundant interleavings.
%
By contrast, {\em stateless} MC (henceforth abbreviated simply as MC)
stores only the current sequence of execution events to avoid a prohibitive memory footprint.
Reduction algorithms \cite{dpor,optimal-dpor,satcheck,mcr,mcr,tsopso}
can then analyze the memory accesses in that sequence to identify
interleavings observationally-equivalent under Mazurkiewicz trace theory \cite{mazurkiewicz} and hence safe to skip.
The resulting state spaces are still exponentially-sized,
but only in the number of conflicting operations rather than all operations.
Of these, Landslide uses Dynamic Partial Order Reduction (DPOR) \cite{dpor} to prune its state spaces.

@@

This section presents our formalization of transactional memory in Landslide's framework of thread concurrency.
We make two major simplifications:
simulating transaction aborts as failure injections,
and treating transaction atomicity as a global mutex during data-race analysis;
and provide corresponding equivalence proofs.

{\bf Notation.} Let $I = TN_1@L_1, TN_2@L_2, ... TN_n@L_n$,
with $N_i$ a thread ID and $L_i$ a code line number,
denote the execution sequence of a program as it runs according to the specified thread interleaving.%
\footnote{This serialization of concurrent execution is told from the perspective of main memory
and hence assumes sequential consistency.
For discussion of relaxed memory models refer to Section~\ref{sec:warpzone}.}

@@

If these instructions from the {\tt x++} in the transaction are preempted,
with another thread's access to {\tt x} interleaved in between,
the transaction will abort.
So, the interleaving
\[
	\tiat{1}, \tiat{2a}, \tiat{2b}, \tjat{1}, \tjat{2}, \tjat{3}, \tiat{2c}, \tiat{3}
\]
or, henceforth abbreviated for clarity:
\[
	\tiat{1-2b}, \tjat{1-3}, \tiat{2c-3}
\]
is not possible; rather, \ti will fall into the backup path:
\[
	\tiat{1-2b}, \tjat{1-3}, \tiat{4-7}
\]
However, the {\tt x++} operation from the failure path (correspondingly $6a$, $6b$, $6c$)
{\em can} be thusly separated with conflicting accesses interleaved in between,
since the mutex only protects the failure path against other failure paths,
but not against the transaction itself.
So (assuming {\tt x} is intended to be a precise counter rather than a sloppy one),
%losing one of the increments to which constitutes a bug),
we observe a bug in the following interleaving.%
\footnote{Note also that this bug requires either at least 3 threads or at least 2 iterations between 2 threads to expose;
this highlights MC's dependence on its test cases to produce meaningful state spaces in the first place.}
\[
	\tiat{1-2b}, \tjat{1-3}, \tiat{4-6b}, \tkat{1-3}, \tiat{6c-7}
\]
Prior work \cite{htm-mario} proposed the idiom shown in Figure~\ref{fig:htm-fixed}
to exclude this family of interleavings,
showing that correctly synchronizing even the simplest transactions may be surprisingly difficult or complex.
%further motivating our research.

@@

Left unstated in interleavings such as $\tiat{1-2c}, \tjat{1-3}, \tiat{4-7}$%
\footnote{For the sake of example I reorder \ti's write to {\tt x} before \tj's part here.}
are HTM's execution semantics, namely:
\begin{enumerate}
	\item any modifications to shared state (such as $2c$) by \ti are not visible to \tj during its execution,
		despite \tj being executed afterwards, and
	\item all local and global state changes by \ti between lines $1$ and $2c$ are discarded when jumping to line $4$.
\end{enumerate}
% there shouldbe a better word than 'requires' here.. benefits from.. "the point of it is"..
While use of TSX in production requires the performance advantage
of temporarily staging such accesses in local CPU cache,
model checking such programs need be concerned only with the program's {\em observable} behaviours.
We claim that MCing the interleaving $\tiat{1}, \tjat{1-3}, \tiat{4-7}$
is an equivalent verification as MCing the one above;
in fact, this interleaving suffices to check
all observable behaviours
of all interleavings
of all subsets of $\tjat{1-3}$
with all subsets of $\tiat{2a-2c}$,
whether they share a memory conflict or not.
Stated formally:

@@

Next, we address the memory accesses within transactions with regard to data-race analysis.
From Theorem~\ref{thm:atom} we have that the body of all transactions may be executed atomically within the MC environment.
While they may interleave within other non-transactional sequences,
no other operations (whether transactional or not) will interrupt them.
We claim this level of atomicity is equivalent to that provided by a global lock,
and hence abstracting it as such in Landslide's data-race analysis is sound.

@@

\begin{itemize}
	\item {\bf Limited Happens-Before.}
		Under this definition, to be a data race in $I$ they must be reorderable at instruction granularity
		\cite{tsan,hybriddatarace}.
		\begin{itemize}
			\llitem $I \Rightarrow I'$:
				If $\tiiat{\mu},\tjjat{\nu}$ race in $I$,
				then they cannot both be in successful transactions,
				or else placing \tiiat{\mu} within the boundaries of \tjjat{\nu}'s transaction
				would cause the latter to abort, invalidating \tjjat{\nu}, or vice versa.
				Hence they will not both hold $m$ in $I'$.
				Otherwise their lock-sets and DPOR dependence relation remain unchanged.
			\llitem $I' \Rightarrow I$:
				If $\tiiat{\mu},\tjjat{\nu}$ race in $I'$,
				both corresponding threads cannot hold $m$;
				WLOG let $\tii$ not hold $m$ during $\tiiat{\mu}$.
				Then in $I$, $\tiiat{\mu}$ is not in a transaction.
				With the remainder of their lock-sets still disjoint,
				and still not DPOR-dependent, $\tjjat{\nu}$ (or its containing transaction)
				can then be reordered directly before or after $\tiiat{\mu}$.
		\end{itemize}
	\item {\bf Pure Happens-Before.}
		WLOG fix $\tiiat{\mu} \prec \tjjat{\nu} \in I$.
		Then under this definition, to be a data race in $I$ there must be no pair of synchronizing instructions
		$\tiiat{\epsilon}$ (a release edge) and $\tjjat{\chi}$ (an acquire edge) such that
		\[
			\tiiat{\mu} \prec \tiiat{\epsilon} \prec \tjjat{\chi} \prec \tjjat{\nu} \in I
		\]
		to update the vector clock epoch between \tiiat{\mu} and \tjjat{\nu} \cite{djit,fasttrack}.
		\begin{itemize}
			\llitem $I \Rightarrow I'$:
				If $\tiiat{\mu},\tjjat{\nu}$ race in $I$,
				then they cannot both be in successful transactions,
				or else Lemma~\ref{lem:equiv} normalization would provide
				the corresponding HTM end and begin for $\tiiat{\epsilon}$ and $\tjjat{\chi}$ respectively.
				Hence there will be no unlock/lock pair on $m$ in $I'$ to satisfy the above sequence.
			\llitem $I' \Rightarrow I$:
				If $\tiiat{\mu},\tjjat{\nu}$ race in $I'$,
				then they cannot both hold $m$,
				or else $\mathsf{lockify}_m$ would provide the corresponding
				unlock and lock for $\tiiat{\epsilon}$ and $\tjjat{\chi}$ respectively.
				Hence there will be no HTM end/begin pair in $I$ to satisfy the above sequence.
		\end{itemize}
		%In both cases $I$ and $I'$ otherwise have the same synchronizing instructions.
\end{itemize}
Hence, data-race analysis is sound when transaction boundaries are replaced by an abstract global lock.
\end{proof}

@@

			& 2,4	&\cpu{*3259.37}&\wtm{436.50}&\ints{*1639}&\ETA{--}&\ETA{>30m}&\ETA{--}& \ETA{36019973} \\

@@

To the best of our knowledge, this is the first work to test transactional programs in a model-checking environment,
%To the best of the author's knowledge, Landslide is the first MC to check transactional concurrency,
so no other MC State-of-the-Art%
\footnote{The author's DJ name.}
exists to compare to in controlled experiments.
Nevertheless, we pose the following evaluation questions.

@@

		{\bf Finding bugs quickly.}
		The multiplicative factor in bug-finding speed (2-4x, eyeballing) is generally smaller
		than that of the total number of interleavings (10-100x)
		as the test parameters increase.
		In other words,
		should they exist,
		Landslide find bugs reasonably quickly in these transactional programs
		despite prohibitive exponential explosion in total state space size.
		This corroborates the prior work \cite{quicksand},
		extending its good news to the world of HTM.
	%\item

@@

\begin{figure}[t]
	%\begin{center}
		\begin{tabular}{l}
		%\texttt{void count() \{} \\
		%\texttt{~~~~for (int i = 0; i < 1000; i++) \{} \\
		\texttt{\flow{while} (\_retry);} \\
		\texttt{\flow{if} (\call{\_xbegin}() == \const{SUCCESS}) \{} \\
		\texttt{~~~~\hilight{brickred}{tie}(\_root,inserted) = \call{\_insert}(\_root,n); } \\
		\texttt{~~~~\call{\_xend}();} \\
		\texttt{\} \flow{else} \{} \\
		\texttt{~~~~\call{pthread\_mutex\_lock}(\&\_tree\_lock);} \\
		\texttt{~~~~\_retry = \const{true};} \\
		\texttt{~~~~\hilight{brickred}{tie}(\_root,inserted) = \call{\_insert}(\_root,n); } \\
		\texttt{~~~~\_retry = \const{false};} \\
		\texttt{~~~~\call{pthread\_mutex\_unlock}(\&\_tree\_lock);} \\
		\texttt{\}} \\
		\end{tabular}
	%\end{center}
	%\caption{mario man had a very large house. or a castle.}
	\caption{Unmodified code from {\tt htmavl.cpp} with the previously-unknown bug found by Landslide.
		The transaction path fails to check {\tt \_retry},
		leading to data races and corruption just as in {\tt htm1}.
		}
	\label{fig:avlbug}
\end{figure}
	%\item Despite Quicksand's ability to optimize for finding bugs in overall smaller state spaces,
	%	this does not correlate with better CPU-time or even better wall-clock time.
	%	While possibly a consequence of the modest test size,
	%	this suggests future work should extend the Quicksand algorithm
	%	to prioritize by state space maximality as well as by state space estimate.
	%\item

@@

\newcommand\ETAdag[1]{\ETA{\ensuremath{\dagger}#1}}
\begin{table}[h]
	\begin{center}
		\begin{tabular}{cc|r|r}
			& & \cpu{\bf cpu (s)} & \ints{\bf SS size} \\
			\bf test & \bf params & \em (or \ETAdag{\bf \em ETA}) & \em (or \ETAdag{\bf \em est.}) \\
			\hline
			\hline
			{\tt htm2}
			& 2,1 & \cpu{18.57}	& \ints{294}	\\
			& 2,2 & \cpu{133.78}	& \ints{4902}	\\
			& 2,3 &\cpu{1986.98}	& \ints{79017}	\\
			& 3,1 &\cpu{11672.15}	& \ints{467730}	\\
			%& 3,2 &	&	\\
			\hline
			{\tt counter}
			& 2,1 & \cpu{5.57}	& \ints{30}	\\
			& 2,2 & \cpu{15.53}	& \ints{384}	\\
			& 2,3 &\cpu{20.00}	& \ints{5280}	\\
			& 2,4 &\cpu{2211.10}	& \ints{75264}	\\
			& 3,1 & \cpu{57.90}	& \ints{1960}	\\
			%& 3,2 &		& 	\\
			\hline
			{\tt swap}
			%& 2,1 & \cpu{38.77}	& \ints{228}	\\ % pre-ignore_thr_func fix
			& 2,1 & \cpu{32.98}	& \ints{193}	\\
			& 2,2 & \cpu{3652.91}	& \ints{101150}	\\
			%[JOB 8] progress: 97913/161897 brs (60.478362%), ETA 1h 29m 30s (elapsed 59m 52s)
			% [JOB 8] TIMED OUT (60.478362%; ETA 1h 29m 30s)

@@

We conclude that MC performance evaluations must address
experimental environment variables
%at least enough
to ensure consistent performance between runs.
So doing, multiple samples and error bars are then necessary
only when using nondeterministic search ordering strategies
such as Iterative Deepening.
Otherwise, considering the low variance shown in Table~\ref{tab:variants},
they need be shown only for a token small test
to promise the reader the same consistency in the larger tests
(especially if the time tradeoff to measure them would sacrifice testing larger state spaces to begin with).
% We also advise the busy researcher to torrent their anime rather than stream it. % lol can't put this in a paper
%% i'm not actually sure this fits..
%Finally, considering recent news on cross-VM cache attacks \cite{meltdown,spectre},
%we also conjecture that cloud-based virtual machines cannot be trusted for low-variance performance measurements.

Outside of this experiment, we fixed our environment by measuring all performance numbers
with Firefox and Chrome as the only other significant machine load.
%(we gotta stay amused somehow).
We believe the exponential differences among completion times justifies the absence of error bars,
which would be expected to show 2\% variance if extrapolating from this experiment.
The numbers of interleavings in each state space are, of course, deterministic and do not vary across runs.

@@

Because Landslide's concurrency model includes only instruction-level thread nondeterminism,
not per-CPU memory buffer reorderings,
our current HTM implementation in Landslide cannot find this bug.
In fact, it erroneously verifies the corresponding test {\tt htm2} in 7 CPU-minutes,
with 294 distinct interleavings in total,
none of which include the above-listed sequence.
Recent work extended DPOR to support TSO and PSO memory nondeterminism \cite{tsopso};
if incorporated into Landslide, we could find or verify the absence of such bugs.
Visual inspection of \cite{htm-mario}'s HTM data structures found no barriers used in this implementation pattern;
we would urge any reader interested in using them to add them in by hand first.

>>>>

\begin{abstract}
Hardware transactional memory is a recently-introduced concurrent programming paradigm
%backed by a set of new instructions on recent x86 CPUs,
which allows programmers to elide locks for performance in low-contention workloads.
However,
%using this feature
it
comes at a cost in implementation complexity:
%programmers must support their
fast-path code must be accompanied by backup paths to handle transaction failure.
%
We extend Landslide, a popular stateless model checker,
with a concurrency model for transactional memory
and evaluate it on several real-world transactional
benchmarks and data structure implementations.
\end{abstract}

@@

Transactional Synchronization Extensions (TSX) \cite{transactional-memory}
is an instruction set extension for x86 CPUs which adds hardware-based transactional memory.
The processor uses its existing cache coherence algorithm to
check for memory conflicts with other cores while temporarily staging a sequence of memory accesses.
If no other CPU accesses the same memory during the transaction,
the access sequence is committed to main memory atomically (with respect to visibility by other CPUs).
Otherwise, the accesses are discarded, the CPU's local state is reverted, and the transaction returns a failure code.

This feature can be used to replace conventional locking in performance-critical concurrent programs.
When a concurrent workload accesses largely thread-local data,
or disjoint sections of a shared data structure,
the contention rate between threads is low,
and transactions will often succeed.
Compared to programs which use conventional locks,
which use bus-locking atomic accesses even in the fastest code path,
TSX provides substantial performance improvements in such programs \cite{htm-experience, htm-performance, htm-mario}.
However, the possibility of transaction failure introduces additional implementation complexity:
programmers must also provide a backup plan
to safely resolve contention between threads,
usually involving conventional synchronization.
These backup paths must coordinate not only with other backup paths
but also with other fast paths which another thread may begin after the original transaction failed,
which even in the simplest transactions requires complex synchronization sequences
\cite{htm-mario}.
This introduces an additional dimension of nondeterminism into an already concurrent program,
and moreover, because transaction failure is expected to be rare,
% だからこそ
obscure interleavings between failure paths are difficult to expose during stress testing.

@@

{\bf Hardware transactional memory.}
TSX was implemented on consumer hardware for the first time by Intel's Haswell architecture \cite{htm-haswell},
which extends the x86 instruction set to provide
{\tt xbegin}, {\tt xend}, and {\tt xabort} for beginning, committing, and aborting
transactions, respectively.
Higher-level programming languages or compilers may offer libraries or intrinsics to access these instructions;
for C and C++ GCC provides intrinsics named {\tt \_xbegin()} and so on \cite{htm-gcc}.
Figure~\ref{fig:htm-example} shows an example program using TSX to synchronize access to a shared counter,
including a failure path which defaults to a conventional lock.
This example actually has a bug, which we will discuss in the next section;
the reader is encouraged to try to spot it before then.
Several related works formally prove the correctness of transactional memory {\em implementations}
\cite{specifying-verifying-tm,tm-correctness,tm-completeness,mc-tm-with-spin},
but verifying the client programs written to use transactions remains an open problem.

@@

Under software transactional memory (STM) \cite{stm-pldi06},
memory conflicts with other threads are the only reason for transaction failure
(apart from programmer-supplied explicit aborts);
hence, depending on program semantics, some transactions may be guaranteed to succeed.
However, hardware transactions (HTM) may also fail
for several other reasons such as random system interrupts or exhausting the CPU's cache capacity.
Because timer interrupts can in principle occur at any moment,
and with arbitrary frequency (observable by the program, perhaps as a result of a heavily-loaded system),
in this paper we will simplify the failure model by saying that HTM transactions can fail for any reason.
We defer discussion of programs which distinguish the reason for aborts through the failure code to Section~\ref{sec:warpzone}.

{\bf Stateless model checking.}
Model checking (MC) \cite{verisoft} is a testing technique for systematically executing and verifying
the possible thread interleavings of a concurrent program.
The main research challenge is to cope with exponential explosion of the state space,
which is sized $O(n^k)$ for a program with $n$ operations and $k$ threads.
Some {\em stateful} MCs explicitly store and compare all visited states of the program being tested \cite{spin},
which both keeps track of test coverage
and allows identifying identical states to avoid testing redundant interleavings.
%
By contrast, {\em stateless} MC (henceforth abbreviated simply as MC)
stores only the current sequence of execution events to avoid a prohibitive memory footprint.
Reduction algorithms \cite{dpor,optimal-dpor,satcheck,mcr,mcr,tsopso}
can then analyze the memory accesses in that sequence to identify
interleavings observationally-equivalent under Mazurkiewicz trace theory \cite{mazurkiewicz} and hence safe to skip.
The resulting state spaces are still exponentially-sized,
but only in the number of conflicting operations rather than all operations.
Of these, Landslide uses Dynamic Partial Order Reduction (DPOR) \cite{dpor} to prune its state spaces.

@@

@@

This section presents our formalization of transactional memory in Landslide's framework of thread concurrency.
We make two major simplifications:
simulating transaction aborts as immediate failure injections,
and treating transaction atomicity as a global mutex during data-race analysis;
and provide corresponding equivalence proofs.

{\bf Notation.} Let $I = TN_1@L_1, TN_2@L_2, ... TN_n@L_n$,
with $N_i$ a thread ID and $L_i$ a code line number,
denote the execution sequence of a program as it runs according to the specified thread interleaving.%
\footnote{This serialization of concurrent execution is told from the perspective of all CPUs at once
and hence assumes sequential consistency.
For discussion of relaxed memory models refer to Section~\ref{sec:warpzone}.}

@@

If these instructions from the {\tt x++} in the transaction are preempted,
with another thread's access to {\tt x} interleaved in between,
the transaction will abort.
So, the interleaving
\[
	\tiat{1}, \tiat{2a}, \tiat{2b}, \tjat{1}, \tjat{2}, \tjat{3}, \tiat{2c}, \tiat{3}
\]
or, henceforth abbreviated for clarity:
\[
	\tiat{1-2b}, \tjat{1-3}, \tiat{2c-3}
\]
is not possible; rather, \ti will fall into the backup path:
\[
	\tiat{1-2b}, \tjat{1-3}, \tiat{4-7}
\]
However, the {\tt x++} operation from the failure path (correspondingly $6a$, $6b$, $6c$)
{\em can} be thusly separated with conflicting accesses interleaved in between,
since the mutex only protects the failure path against other failure paths,
but not against the transaction itself.
So (assuming {\tt x} is intended to be a precise counter rather than a sloppy one),
%losing one of the increments to which constitutes a bug),
we observe a bug in the following interleaving.%
\footnote{Note also that this bug requires either at least 3 threads or at least 2 iterations between 2 threads to expose;
this highlights MC's dependence on its test cases to produce meaningful state spaces in the first place.}
\[
	\tiat{1-2b}, \tjat{1-3}, \tiat{4-6b}, \tkat{1-3}, \tiat{6c-7}
\]
Prior work \cite{htm-mario} proposed the idiom shown in Figure~\ref{fig:htm-fixed}
to exclude this family of interleavings,
which shows that correctly synchronizing even the simplest transactions may be surprisingly difficult or complex.
%further motivating our research.

@@

Left unstated in interleavings such as $\tiat{1-2c}, \tjat{1-3}, \tiat{4-7}$%
\footnote{For a clearer example we reorder \ti's write to {\tt x} before \tj's part here.}
are HTM's execution semantics, namely:
\begin{enumerate}
	\item any modifications to shared state (such as $2c$) by \ti are not visible to \tj during its execution,
		despite \tj being executed afterwards, and
	\item all local and global state changes by \ti between lines $1$ and $2c$ are discarded when jumping to line $4$.
\end{enumerate}
% there shouldbe a better word than 'requires' here.. benefits from.. "the point of it is"..
While use of TSX in production requires the performance advantage
of temporarily staging such accesses in local CPU cache,
model checking such programs need be concerned only with the program's {\em observable} behaviours.
We claim that MCing the simpler interleaving $\tiat{1}, \tjat{1-3}, \tiat{4-7}$
is an equivalent verification as MCing the one above;
in fact, this interleaving suffices to check
all observable behaviours
of all interleavings
of all subsets of $\tjat{1-3}$
with all subsets of $\tiat{2a-2c}$,
whether they share a memory conflict or not.
Stated formally:

@@

Next, we address the memory accesses within transactions with regard to data-race analysis.
From Theorem~\ref{thm:atom} we have that the body of all transactions may be executed atomically within the MC environment.
While they may interleave between other non-transactional sequences,
no other operations (whether transactional or not) will interrupt them.
We claim this level of atomicity is equivalent to that provided by a global lock,
and hence abstracting it as such in Landslide's data-race analysis is sound.

@@

\begin{itemize}
	\item {\bf Limited Happens-Before.}
		%Under this definition,
		To race in $I$ they must be reorderable at instruction granularity,
		at least one with a thread switch immediately before or after.
		\cite{tsan,hybriddatarace}.
		\begin{itemize}
			\llitem $I \Rightarrow I'$:
				If $\tiiat{\mu},\tjjat{\nu}$ race in $I$,
				then they cannot both be in successful transactions,
				or else placing \tiiat{\mu} within the boundaries of \tjjat{\nu}'s transaction
				would cause the latter to abort, invalidating \tjjat{\nu}, or vice versa.
				Hence they will not both hold $m$ in $I'$.
				Otherwise their lock-sets and DPOR dependence relation remain unchanged.
			\llitem $I' \Rightarrow I$:
				If $\tiiat{\mu},\tjjat{\nu}$ race in $I'$,
				both corresponding threads cannot hold $m$;
				WLOG let $\tii$ not hold $m$ during $\tiiat{\mu}$.
				Then in $I$, $\tiiat{\mu}$ is not in a transaction.
				With the remainder of their lock-sets still disjoint,
				and still not DPOR-dependent, $\tjjat{\nu}$ (or its containing transaction)
				can then be reordered directly before or after $\tiiat{\mu}$.
		\end{itemize}
	\item {\bf Pure Happens-Before.}
		WLOG fix $\tiiat{\mu} \prec \tjjat{\nu} \in I$.
		Then to race in $I$ there must be no pair of synchronizing instructions
		$\tiiat{\epsilon}$ (a release edge) and $\tjjat{\chi}$ (an acquire edge) such that
		\[
			\tiiat{\mu} \prec \tiiat{\epsilon} \prec \tjjat{\chi} \prec \tjjat{\nu} \in I
		\]
		to update the vector clock epoch between \tiiat{\mu} and \tjjat{\nu} \cite{djit,fasttrack}.
		\begin{itemize}
			\llitem $I \Rightarrow I'$:
				If $\tiiat{\mu},\tjjat{\nu}$ race in $I$,
				then they cannot both be in successful transactions,
				or else Lemma~\ref{lem:equiv} normalization would provide
				the corresponding HTM end and begin for $\tiiat{\epsilon}$ and $\tjjat{\chi}$ respectively.
				Hence there will be no unlock/lock pair on $m$ in $I'$ to satisfy the above sequence.
			\llitem $I' \Rightarrow I$:
				If $\tiiat{\mu},\tjjat{\nu}$ race in $I'$,
				then they cannot both hold $m$,
				or else $\mathsf{lockify}_m$ would provide the corresponding
				unlock and lock for $\tiiat{\epsilon}$ and $\tjjat{\chi}$ respectively.
				Hence there will be no HTM end/begin pair in $I$ to satisfy the above sequence.
		\end{itemize}
		%In both cases $I$ and $I'$ otherwise have the same synchronizing instructions.
\end{itemize}
Hence, data-race analysis is sound when transaction boundaries are replaced by an abstract global lock.
\end{proof}

@@

			& 2,4	&\cpu{*3259.37}&\wtm{*436.50}&\ints{*1639}&\ETA{--}&\ETA{>30m}&\ETA{--}& \ETA{36019973} \\

@@

To the best of our knowledge, this is the first work to test transactional programs in a model-checking environment,
%To the best of the author's knowledge, Landslide is the first MC to check transactional concurrency,
so no other MC State of the Art%
\footnote{The author's DJ name.}
exists to compare to in controlled experiments.
Nevertheless, we pose the following evaluation questions.

@@

		{\bf Finding bugs quickly.}
		As the test parameters increase,
		the multiplicative factor in bug-finding speed (2-4x, eyeballing) is generally smaller
		than that of the total number of interleavings (10-100x).
		In other words,
		should they exist,
		Landslide find bugs reasonably quickly in these transactional programs
		despite prohibitive exponential explosion in total state space size.
		This corroborates the prior work \cite{quicksand},
		extending its good news to the world of HTM.
	%\item

@@

\begin{figure}[t]
	%\begin{center}
		\begin{tabular}{l}
		%\texttt{void count() \{} \\
		%\texttt{~~~~for (int i = 0; i < 1000; i++) \{} \\
		\texttt{\flow{while} (\_retry);} \\
		\texttt{\flow{if} (\call{\_xbegin}() == \const{SUCCESS}) \{} \\
		\texttt{~~~~\hilight{brickred}{tie}(\_root,inserted) = \call{\_insert}(\_root,n); } \\
		\texttt{~~~~\call{\_xend}();} \\
		\texttt{\} \flow{else} \{} \\
		\texttt{~~~~\call{pthread\_mutex\_lock}(\&\_tree\_lock);} \\
		\texttt{~~~~\_retry = \const{true};} \\
		\texttt{~~~~\hilight{brickred}{tie}(\_root,inserted) = \call{\_insert}(\_root,n); } \\
		\texttt{~~~~\_retry = \const{false};} \\
		\texttt{~~~~\call{pthread\_mutex\_unlock}(\&\_tree\_lock);} \\
		\texttt{\}} \\
		\end{tabular}
	%\end{center}
	%\caption{mario man had a very large house. or a castle.}
	\caption{Unmodified code from {\tt htmavl.cpp} showing the previously-unknown bug found by Landslide.
		The transaction path fails to check {\tt \_retry},
		leading to data races and corruption just as in {\tt htm1}.
		}
	\label{fig:avlbug}
\end{figure}
	%\item Despite Quicksand's ability to optimize for finding bugs in overall smaller state spaces,
	%	this does not correlate with better CPU-time or even better wall-clock time.
	%	While possibly a consequence of the modest test size,
	%	this suggests future work should extend the Quicksand algorithm
	%	to prioritize by state space maximality as well as by state space estimate.
	%\item

@@

\newcommand\ETAdag[1]{\ETA{\ensuremath{\dagger}#1}}
\begin{table}[h]
	\begin{center}
		\begin{tabular}{cc|r|r}
			& & \cpu{\bf cpu (s)} & \ints{\bf SS size} \\
			\bf test & \bf params & \em (or \ETAdag{\bf \em ETA}) & \em (or \ETAdag{\bf \em est.}) \\
			\hline
			\hline
			{\tt htm2}
			& 2,1 & \cpu{18.57}	& \ints{294}	\\
			& 2,2 & \cpu{133.78}	& \ints{4902}	\\
			& 2,3 &\cpu{1986.98}	& \ints{79017}	\\
			& 3,1 &\cpu{11672.15}	& \ints{467730}	\\
			& 3,2 &	&	\\ % TODO currently running
			\hline
			{\tt counter}
			& 2,1 & \cpu{5.57}	& \ints{30}	\\
			& 2,2 & \cpu{15.53}	& \ints{384}	\\
			& 2,3 &\cpu{20.00}	& \ints{5280}	\\
			& 2,4 &\cpu{2211.10}	& \ints{75264}	\\
			& 3,1 & \cpu{57.90}	& \ints{1960}	\\
			& 3,2 &		& 	\\ % TODO
			\hline
			{\tt swap}
			%& 2,1 & \cpu{38.77}	& \ints{228}	\\ % pre-ignore_thr_func fix
			& 2,1 & \cpu{32.98}	& \ints{193}	\\
			& 2,2 & \cpu{3652.91}	& \ints{101150}	\\
			%[JOB 8] progress: 97913/161897 brs (60.478362%), ETA 1h 29m 30s (elapsed 59m 52s)
			% [JOB 8] TIMED OUT (60.478362%; ETA 1h 29m 30s)

@@

We conclude that MC performance evaluations must address
experimental environment variables
%at least enough
to ensure consistent performance between runs.
So doing, multiple samples and error bars are then necessary
only when using nondeterministic search ordering strategies
such as Iterative Deepening.
Otherwise, considering the low variance shown in Table~\ref{tab:variants},
they need be shown only for a token small test
to provide reader some assurance of similar consistency in the larger tests
(especially if the time tradeoff to measure them would sacrifice testing larger state spaces to begin with).
% We also advise the busy researcher to torrent their anime rather than stream it. % lol can't put this in a paper
%% i'm not actually sure this fits..
%Finally, considering recent news on cross-VM cache attacks \cite{meltdown,spectre},
%we also conjecture that cloud-based virtual machines cannot be trusted for low-variance performance measurements.

%Outside of this experiment,
For all tests outside of Table~\ref{tab:variants},
we fixed our environment by measuring all performance numbers
with Firefox and Chrome as the only other significant machine load.
%(we gotta stay amused somehow).
We believe the exponential differences among completion times justifies the absence of error bars,
which one would expect to show 2\% variance if extrapolating from these results. %this experiment.
The numbers of interleavings in each state space are, of course, deterministic and do not vary across runs.

@@

Because Landslide's concurrency model includes only instruction-level thread nondeterminism,
not per-CPU memory buffer reorderings,
our current HTM implementation cannot find this bug.
In fact, it erroneously verifies the corresponding test {\tt htm2}(3,1)
% lol this is both old (quicksand) AND wrong (testing (2,1) which can't expose the bug)
%in 7 CPU-minutes,
%with 294 distinct interleavings in total,
in 3 CPU-hours,
with 467730 distinct interleavings in total,
none of which include the above-listed sequence.
Recent work extended DPOR to support TSO and PSO memory nondeterminism \cite{tsopso};
if incorporated into Landslide, we could find or verify the absence of such bugs.
Visual inspection of \cite{htm-mario}'s HTM data structures found no barriers used in this implementation pattern;
we would urge any reader interested in using those to add them in by hand first.
